---
title: "Notebook for Regression Discontinuity Design (RDD)"
author:
- name: Aleli D. Kraft, PhD

date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: lumen
    code_folding: show
    toc: yes
    toc_float: yes
    toc_depth: 4
    number_sections: no
    highlight: pygments
    fig_caption: yes
editor_options:
  markdown:
    wrap: 72
---

# Codes for Regression Discontinuity Design

```{r, echo = FALSE, warning = FALSE, message = FALSE}
packages <- c("tidyverse", "haven", "gtsummary", "jtools", "ggplot2", "rdrobust", "rddensity", "here");for(pack in packages){
  if(!require(pack, character.only = TRUE, quietly = TRUE)){
    install.packages(pack, dependencies = TRUE)
    require(pack, character.only = TRUE)
  }
  # Clean up the workspace
  rm(pack)
};theme_set(theme_apa())

```

# Slide 3
## Check whether there is any evidence of manipulation of the eligibility index 

```{r}
#Load dataset
hisp_data <- read_dta(here("evaluation.dta"))
hisp_data <- hisp_data %>%
  mutate(D = ifelse(poverty_index > 58, "> 58", "<= 58"))

#Check whether there is any evidence of manipulation of the eligibility index 
ggplot(hisp_data, aes(x = poverty_index)) + 
  geom_vline(xintercept = 58, linetype = 'dotted') +
  geom_density()+
  ggtitle("Density Plot of Poverty index")


```


# Slide 4
## Can you apply a sharp RDD? 

```{r, warning = FALSE}
#Categorize > 58 and <= 58
hisp_data_treatment <- hisp_data %>%
  filter(treatment_locality == 1) 
#Can you apply a sharp RDD? 
rdplot(y = hisp_data_treatment$enrolled,
       x = hisp_data_treatment$poverty_index,
       c = 58,
       p = 1,
       nbins = c(58, 42),
       x.label = "Forcing variable (Poverty index)",
       y.label = "Percent of households that participate (Enrolled)",
       )

# Plotting
ggplot(hisp_data_treatment, aes(y = enrolled, x = poverty_index, color = factor(D))) +
  geom_vline(xintercept = 58, linetype = "dotted") +
  geom_point() +
  ggtitle("Enrolled vs Poverty Index") + 
  labs(x = "Forcing variable (Poverty index)", y = "Outcome variable (Enrolled)")


```

# Slide 5
## Estimate the sharp RDD estimates. Plot the difference in the health expenditures between the eligible and non-eligible. 

```{r}
# Normalize the poverty index at "0"
hisp_data_model <- hisp_data %>%
  mutate(poverty_index_left = ifelse(poverty_index <= 58, poverty_index - 58, 0),
         poverty_index_right = ifelse(poverty_index > 58, poverty_index - 58, 0)) %>%
  filter(round == 1 & treatment_locality == 1)

# Apply regression on a subset of data
reg_model <- lm(health_expenditures ~ poverty_index_left + poverty_index_right + eligible,
                data = hisp_data_model)

# Extract fitted values
hisp_data_model$he_pred1<- reg_model$fitted.values

# Plotting
ggplot(hisp_data_model, aes(x = poverty_index, y = he_pred1, color = factor(D))) +
    geom_vline(xintercept = 58, linetype = "dotted") +
  ggtitle("RDD Plot") + 
  geom_point() + labs(x = "Forcing variable (Poverty index)", y = "Predicted Health Expenditures")
```

# Slide 6
## Estimate the effect using the rdrobust command

```{r}
rdd <- rdrobust(
  y = hisp_data_model$health_expenditures,
  x = hisp_data_model$poverty_index,
  c = 58
)

summary(rdd)
```

# Slide 7
## Estimate the impact using RDD with controls
Use the following variables as your controls: age_hh age_sp educ_hh educ_sp female_hh indigenous hhsize dirtfloor bathroom land hospital_distance

```{r}
# Define the model formula
model_formula <- health_expenditures ~ eligible + poverty_index_left + poverty_index_right +
  age_hh + age_sp + educ_hh + educ_sp + female_hh + indigenous +
  hhsize + dirtfloor + bathroom + land + hospital_distance

# Filter the data and run the regression
reg_model_cov <- lm(model_formula, data = hisp_data_model)

# Summarize the regression results
summary(reg_model_cov)

```

# Slide 8
## Introduce in your regression model squared, cubic, and quartic poverty score terms.



```{r}
# Generating higher order terms
hisp_data_model <- hisp_data_model %>%
  mutate(
    poverty_index_left2 = poverty_index_left^2,
    poverty_index_right2 = poverty_index_right^2,
    poverty_index_left3 = poverty_index_left^3,
    poverty_index_right3 = poverty_index_right^3,
    poverty_index_left4 = poverty_index_left^4,
    poverty_index_right4 = poverty_index_right^4
  )

# Define the model formula for regression with higher order terms
model_formula <- health_expenditures ~ eligible + poverty_index_left + poverty_index_right +
  poverty_index_left2 + poverty_index_right2 +
  poverty_index_left3 + poverty_index_right3 +
  poverty_index_left4 + poverty_index_right4

# Filter the data and run the regression
reg_model_cov_polynomial <- lm(model_formula, data = hisp_data_model)

# Summarize the regression results
summary(reg_model_cov_polynomial)

```

# Slide 9
## Plot the fitted value of out-of-pocket health expenditures estimated in the last equation against the poverty score. Is there any evidence of non-linearity?


```{r}
######################SLIDE 8####################################
hisp_data_model$he_pred2 <- reg_model_cov_polynomial$fitted.values

# Plotting
ggplot(hisp_data_model, aes(x = poverty_index, y = he_pred2, color = factor(D))) +
  geom_point() + geom_vline(xintercept = 58, linetype = "dotted") + labs(x = "Forcing variable (Poverty index)", y = "Predicted Health Expenditures")

```

# Slide 10
## Perform a “false experiment” using the RDD model in the baseline period.
Plot the fitted value of out-of-pocket health expenditures estimated in the last step against the poverty score. 


```{r}
# Filter the data
filtered_data <- filter(hisp_data, round == 0 & treatment_locality == 1)

# Regression model
reg_model_false <- lm(health_expenditures ~ eligible + poverty_index, data = filtered_data)

# Generate predictions
filtered_data$he_pred3 <- reg_model_false$fitted.values

# Plotting the predictions
ggplot(filtered_data, aes(x = poverty_index, y = he_pred3, color = factor(D))) +
  geom_point() +
  labs(x = "Poverty Index", y = "Predicted Health Expenditures",
       title = "Predicted Health Expenditures vs. Poverty Index")

```

# Slide 10
## Use the rbrobust command to estimate the impact

```{r}
rdd <- rdrobust(
  y = filtered_data$health_expenditures,
  x = filtered_data$poverty_index,
  c = 58
)

summary(rdd)

```
